---
title: "A brief Introduction to R"
author: "Song Wang"
date: "January 20, 2016"
output: html_document
---

### 1. Download and install R and Rstudio
Visit http://cran.us.r-project.org

https://www.rstudio.com/products/rstudio/download/

### 2. play with R 
- arithmetic operation 
```{r, echo= TRUE, eval=TRUE}
1+2
3*4
3/6
3>5
```
- use easy in-built functions 
```{r}
x <- c(12, 11, 16, 20)  # c standards for ?
?c
sum(x) 
max(x)
mean(x)
median(x)
sd(x)
var(x)

?mean
```

### 3. Get Data into R

- use import Dataset button in the top right cornor of Rstudio (discouraged)
 click the button, select the file you want to import.

- use read.table command
```{r}
rm(list=ls())
mtcars <- read.table(file = "mtcars.csv", sep=",", header = T)  
# mtcars <- read.csv(file = "mtcars.csv")
?read.csv
?read.table
str(mtcars)
?mtcars
mean(mtcars$mpg)
```




### 4. Visualize your data 
```{r}
# for numeric variable (1 or 2 variables)

hist(mtcars$mpg)
hist(mtcars$mpg, freq = FALSE) #density histogram

plot(density(mtcars$mpg))
plot(density(mtcars$disp))

boxplot(mtcars$mpg, horizontal = T)
stripchart(mtcars$mpg,method = 'jitter')


# two numerical variables
plot(mtcars$disp, mtcars$mpg, xlab ="displacement", ylab = "mile per gallon", main = "mpg vs. disp") # scatter plot
#abline(lm( mtcars$mpg ~ mtcars$disp ), col="red")

pairs(mtcars[,1:4]) # pairwise plot

# For categorial data 
counts = table(mtcars$cyl); barplot(counts)
boxplot(mtcars$mpg ~ mtcars$cyl)
boxplot(mtcars$hp ~ mtcars$cyl)
```


- Checking Normality 
```{r}
hist(mtcars$mpg)
qqnorm(mtcars$mpg)
qqline(mtcars$mpg)

set.seed(123)
x <- rnorm(100,1,2)
hist(x)
qqnorm(x)
qqline(x)
```


### 5. How to save the plot and use in the report later
- click on the plot (at bottom right corner usually) 
- save the plot in .png or .pdf
- copy the plot into the report (in the future project)

  



